FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2023-08-18 paper.tex
abstract, sec 1, sec 2

\Fp{} numbers use a finite number of bits to represent a spectrum of points along the real number line~(\cref{f:real-vs-fp}).
The implementation strategy is essentially scientific notation.
A \fp{} number packs a sign bit, an exponent, and a fraction part (also called the ``significand'' or ``mantissa'') into a bitstring (typically 64 or 32 bits, but 16-bit and 8-bit formats are on the rise~\cite{klowerLowprecisionClimateComputing2021,fp8}).
This representation supports very small and very large numbers in a narrow range of bits.

@@

IEEE~754 defers the question of how to handle such exceptions to application code.
It is up to developers to watch for \Nan{} and \Inf{} (and also subnormal numbers)
and implement an appropriate repair.
Without guards, bugs happen.

@@

Exceptional values have a lifetime: they are born, or \emph{generated}, by some operation; they \emph{propagate} through other operations; and they either appear in the program output, go out of scope, or get \emph{killed} by a numeric operation.
\Cref{f:gpk} summarizes this \emph{gen}-\emph{prop}-\emph{kill} process.
The gens and props are straightforward; see above for examples~(\cref{s:exnvalue}).
The kills often arise from numeric comparisons (\code{<}, \code{=}, etc.),
but exponents (\code{1\^NaN}) and over-eager matrix optimizations can kill
exceptions~\cite{ddghlllprr-correctness-2022}.

To illustrate TODO

Consider the following two functions to find the maximum value in a list:

@@

Both \code{max1} and \code{max2} compute the maximum in a list.
However, the \texttt{<=} operator used in the \code{max1} implementation does \emph{not} propagate \NaN{}s: \texttt{NaN <= 4} returns \texttt{false}, not \NaN{}.
Because of this, you might not get the result you expected:

\begin{lstlisting}[language = Julia]
max1([1, 5, NaN, 4])     # returns 4
max2([1, 5, NaN, 4])    # returns NaN
\end{lstlisting}

Not only is the result from \texttt{max1} problematic for obscuring the fact that there was a \NaN{} in the list, the result is also wrong!
The built-in \texttt{max} function correctly propagates \NaN{}s, so we can see from its return that we got a \NaN{} somewhere.
It's not entirely obvious \emph{where} that \NaN{} came from.

@@

\FlowFPX{} is a set of tools to aid in tracking down \fp{} exceptions.
The two primary tools we cover in this paper are \FT{} and CSTG:

\begin{itemize}
\item \FT{} traces the lifetime of exceptional \fp{} values and enables inside-out fuzzing.
\item CSTG visualizes the tracing output from \FT{} in compact form.
\end{itemize}

\GPUFPX~\cite{llsflg-hpdc-2023} is another component of \FlowFPX{} that can track \fp{} exceptions inside a GPU kernel.
See the cited paper for more details on \GPUFPX.

%% TODO julia proposed consistent NaN handling,
%%  do we propagate the exact NaN
%%  https://github.com/JuliaLang/julia/issues/48523
%%
%% Ashton: @Ben we simply propagate whatever we got from the operation. If
%% Julia's behavior around an operator change, that should be preserved even
%% when using FloatTracker

@@

>>>>

\Fp{} numbers use a finite number of bits to represent a spectrum of points along the real number line~(\cref{f:real-vs-fp}).
The implementation strategy is essentially scientific notation.
A \fp{} number packs a sign bit, an exponent, and a fraction part (also called the ``significand'' or ``mantissa'') into a bitstring.
Typical strings are 64 or 32 bits long, but 16-bit and 8-bit formats are on the rise~\cite{klowerLowprecisionClimateComputing2021,fp8}.
This representation supports very small and very large numbers in a narrow range of bits:

@@

IEEE~754 defers the question of how to handle such exceptions to application code.
It is up to developers to watch for \Nan{}, \Inf{}, and subnormal numbers (underflow)
and implement an appropriate repair.
This is no easy task, however, because of the approximations inherent to
floating point.
Even an apparently-safe division could result in a \Nan{} if its
denominator gets truncated to zero.
Some \Nan{}s might be spurious, others might be fatal, but in any event
anticipating the various exceptions is a burden.
All too often, exceptional values go unhandled and flow through the code.

@@

Unhandled exceptional values have a lifetime: they are born, or \emph{generated}, by some operation; they \emph{propagate} through other operations; and they either appear in the program output, go out of scope, or get \emph{killed} by a numeric operation.
\Cref{f:gpk} summarizes this \emph{gen}-\emph{prop}-\emph{kill} process.
The gens and props are straightforward; see above for examples~(\cref{s:exnvalue}).
The kills often arise from numeric comparisons (\code{<}, \code{=}, etc.),
but exponents (\code{1{\string^}NaN}) and over-eager matrix
optimizations~\cite{ddghlllprr-correctness-2022} can kill exceptions as well.

To illustrate the perils of killed exceptions, consider the following two ways
of finding the maximum value in a list.
The first compares numbers with \code{<=}
while the second uses the built-in \code{max} function:

@@

For lists with a \Nan{} inside, the functions can give different
results because \code{<=} kills \Nan{}s whereas \code{max}
propagates them:

\begin{lstlisting}[language = Julia]
julia> max1([1, 5, NaN, 4]) 
4.0
julia> max2([1, 5, NaN, 4]) 
NaN
\end{lstlisting}

Not only is the result from \code{max1} problematic for obscuring the fact that there was a \NaN{} in the list, the result is arguably wrong!
The result from \code{max2} at least shows that a \Nan{} was in the works, though in a realistic setting it may not be clear where the \Nan{} came from.
Both versions would thus benefit from tools that track exceptions across their lifetime.
FlowFPX can help.

@@

\FlowFPX{} is a toolkit for tracking down \fp{} exceptions.
The two primary tools we cover in this paper are \FT{}, which records lifetimes
and enables fuzzing, and \CSTG{}s (more precisely, CSTGs), which visualize
the flow of exceptions.
\GPUFPX{} is a third component of \FlowFPX{} that tracks \fp{} exceptions inside GPU kernels~\cite{llsflg-hpdc-2023}.

@@

FILL Consistent exception handling.

https://github.com/JuliaLang/julia/issues/48523

Ashton: @Ben we simply propagate whatever we got from the operation. If
Julia's behavior around an operator change, that should be preserved even
when using FloatTracker

Ben: @Ashton tell the JuliaCon readers not just me!
